Nucleotide Diversity Loss on a Plant Y Chromosome Following Recent Recombination Suppression

### 2016-11-13 manuscript/Ydiversity_nov12.tex
minor edits, start on results

\subsection*{Simulations of purifying selection}
To study the effects of purifying selection on expected levels of Y-chromosome diversity, we used a similar approach to \citep{Wilsonsayres2014} and conducted forward-time simulations of haploid Y chromosomes using the software SFSCODE \citep{hernandez2008flexible}. We first estimated the distribution of fitness effects of deleterious mutations from our polymorphism data for X-linked genes using the method of \citep{keightley2007joint}, which fits a gamma distribution of selection coefficients to the observed frequency distribution of nonsynonymous and synonymous polymorphism. We used this estimated gamma distribution to parameterize the simulations, initializing each with our estimated $\theta$ from autosomal genes, but appropriately adjusted to reflect the expected reduction in Y-linked $N_{e}$ for a sex ratio of $N_{f}/(N_{f}+N_{m})=0.6$. To match our sample size and the number of synonymous sites sample from our data (see Supporting Information), the simulations sampled 6 haploid chromosomes, and simulated sequences contained 45,331bp of linked neutral sequence from which we calculated diversity. To examine the expected reduction in diversity relative to neutrality ($\pi/\pi_{0}$) as a function of the number of selected sites ($L$), and to estimate this parameter from our data, we ran simulations over a range of values of $L$ (up to a maximum of $5x10^{6}$) (Figure 3A). For each simulation, we calculated the approximate likelihood of our observed data based on the proportion of simulations in which synonymous diversity, $\pi_{s}_{simulated}$, matched our empirical estimate, $\pi_{s}_{observed}$ (Figure 3B).

\section*{Results and Discussion}

\subsection*{Extensive loss of Y-chromosome diversity}

>>>>

\subsection*{Simulations of purifying selection}
To study the effects of purifying selection on expected levels of Y-chromosome diversity, we conducted forward-time simulations of haploid Y chromosomes using the software SFSCODE \citep{hernandez2008flexible}. We first estimated the distribution of fitness effects of deleterious mutations from our polymorphism data for X-linked genes using the method of \citep{keightley2007joint}, which fits a gamma distribution of selection coefficients to the observed frequency distribution of nonsynonymous and synonymous polymorphism. We used this estimated gamma distribution to parameterize the simulations, initializing each with our estimated $\theta$ from autosomal genes, but appropriately adjusted to reflect the expected reduction in Y-linked $N_{e}$ for a sex ratio of $N_{f}/(N_{f}+N_{m})=0.6$. To match our sample size and the number of synonymous sites sample from our data (see Supporting Information), the simulations sampled 6 haploid chromosomes, and simulated sequences contained 45,331bp of linked neutral sequence from which we calculated diversity. To examine the expected reduction in diversity ($\pi/\pi_{0}$) as a function of the number of selected sites ($L$), and to estimate this parameter from our data, we ran simulations over a range of values of $L$ (up to a maximum of $5x10^{6}$) (Figure 3A). For each such simulation, we calculated the approximate likelihood of our observed data based on the proportion of simulations in which synonymous diversity, $\pi_{s}_{simulated}$, was not significantly different from our empirical estimate, $\pi_{s}_{observed}$ (Figure 3B).

\section*{Results and Discussion}
\subsection*{Extensive loss of Y-chromosome diversity}
%Our analysis of polymorphism levels across the genome in the dioecious plant \textit{R. hastatulus} reveals that neutral diversity on the Y chromosome is substantially lower than expected from a standard neutral model. In particular, neutral diversity on Y-linked genes $\theta_{Y}$ is on average \textasciitilde 2.1\% of the value for their homologues on the X chromosome, and \textasciitilde 2.1\% of the value of autosomal genes. Talking 1/4 of mean autosomal diversity as the equilibrium expectation under neutrality \citep{wright1931evolution}, this corresponds to a chromosome-wide reduction of \textasciitilde 93\% ($\pi/\pi_{0}$). By comparing X and Y diversity estimates to that estimated on autosomal genes, our results indicate that the difference in diversity between X and Y homologues is not due to an elevation of diversity on the X chromosome, but rather a Y-specific reduction. Our results to suggest a slight elevation in X-linked diversity relative to autosomes ($X/A=$), but this difference is not significant different from expectations under equal sex ratios (P), or the empirically-estimated ratio of 0.6 (P). Although our estimates of diversity have not beed normalized by divergence, previous estimates of the average synonymous substitution rate, weighted by the number of synonymous sites, between \textit{R. hastatulus} and the non-dioecious outgroup \textit{R. bucephalophorus} revealed similar levels of synonymous divergence at sex-linked (0.2016) and autosomal genes (0.219) \citep{hough2014}. It is therefore unlikely that our results are caused by mutation rate differences between sex-linked and autosomal genes.

### 2016-11-13 manuscript/manuscript.tex
minor edits, start on results

\subsection*{Extensive loss of Y-chromosome diversity}
Our analysis reveals that diversity on the \textit{R. hastatulus} Y chromosome is significantly lower than expected under neutrality, with estimates indicating Y/A=0.02, ....which is 12.5 fold lower than the standard neutral prediction of Y/A = 0.25 (P<0.0001). We also observe that the Y chromosome shows a 40-fold lower than mean diversity on the X chromosome (Table 1). Note that by normalizing X and Y diversity by autosomal diversity, our results indicate that the X-Y difference we observed was not due to an elevation of X chromosome diversity, but rather a Y-specific reduction. Conceivably, such low diversity on the Y could arise from a low Y-linked mutation rate, or a lower mutation rate in males compared to females. However, these possibilities are unlikely because the number of synonymous mutations in X and Y lineages, estimated by both parsimony and maximum likelihood, are not significantly different \citep{hough2014}. Similarly, our estimates of weighted average synonymous substitution rate between \textit{R. hastatulus} and the outgroup \textit{R. bucephalophorus} reveal similar levels of synonymous divergence at sex-linked (0.2016) and autosomal genes (0.219) \citep{hough2014}.

>>>>

\subsection*{Extensive loss of Y-chromosome diversity}
Our analysis of polymorphism levels across the genome in the dioecious plant \textit{R. hastatulus} reveals that neutral diversity on the Y chromosome is substantially lower than expected from a standard neutral model. In particular, neutral diversity on Y-linked genes $\theta_{Y}$ is on average \textasciitilde 2.1\% of the value for their homologues on the X chromosome, and \textasciitilde 2.1\% of the value of autosomal genes. Talking 1/4 of mean autosomal diversity as the equilibrium expectation under neutrality \citep{wright1931evolution}, this corresponds to a chromosome-wide reduction of \textasciitilde 93\% ($\pi/\pi_{0}$). By comparing X and Y diversity estimates to that estimated on autosomal genes, our results indicate that the difference in diversity between X and Y homologues is not due to an elevation of diversity on the X chromosome, but rather a Y-specific reduction. Our results to suggest a slight elevation in X-linked diversity relative to autosomes ($X/A=$), but this difference is not significant different from expectations under equal sex ratios (P), or the empirically-estimated ratio of 0.6 (P). Although our estimates of diversity have not beed normalized by divergence, previous estimates of the average synonymous substitution rate, weighted by the number of synonymous sites, between \textit{R. hastatulus} and the non-dioecious outgroup \textit{R. bucephalophorus} revealed similar levels of synonymous divergence at sex-linked (0.2016) and autosomal genes (0.219) \citep{hough2014}. It is therefore unlikely that our results are caused by mutation rate differences between sex-linked and autosomal genes.
